---
title: "L13 - ioslides"
author: "Vincent Latona"
date: "`r format(Sys.time(), '%B %d, %Y')`\n"
logo: asulogo.png
output: ioslides_presentation
transition: faster
widescreen: true
smaller: true
---

<style>
.gdbar img{
width: 150px !important;
height: 80px !important;
margin: 8px 8px;
}

.gdbar{
width: 200px !important;
height: 100px !imporant;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, comment="", cache=F)
library(magrittr)
library(leaflet)
library(plotly)
```

## Dataset `trees` {data-background=#00B9E3}
```{r echo=TRUE}
data(trees)
head(trees)
```

## Dataset `trees`; Volume vs Girth {data-background=Panda.jpg data-background-size=cover}
<p style="font-size: 16px"> model: $\text{Volume} = 
\beta_0 + \beta_1\cdot\text{Girth} + \varepsilon; \hspace{1 cm} \varepsilon \sim N(0; \sigma^2)$ <br>
&nbsp; fitted: $\text{Volume} = \hat{\beta_0} + \hat{\beta_1} \cdot 
\text{Girth}$ &nbsp; &nbsp; &nbsp;
&nbsp; &nbsp; &nbsp; &nbsp; $\hat{\beta_0} = b_0 - \text{estimate of }\beta_0$;
$\hat{\beta_1} = b_1 - \text{estimate of }\beta_1$
<\p>

```{r, echo=FALSE}
data(trees)
mod = lm(Volume ~ Girth, data=trees)
x = trees$Girth
y = trees$Volume
xax = list(title="Girth",
           titlefont=list(family="Modern Computer Roman"))
yax = list(title="Volume",
           titlefont=list(family="Modern Computer Roman"))
fig = plot_ly()
```

## Dataset `diamond` from `UsingR`
```{r, message=FALSE}
library(UsingR)
library(plotly)
data(diamond)
head(diamond)
```

## Dataset `mtcars`
```{r}
data(mtcars)
attach(mtcars)
### <b>
head(mtcars)
### </b>
```

## Plot
```{r, message=FALSE}
xax = list(title="Weight",
           titlefont=list(family="Modern Computer Roman"))
yax = list(title="Miles per Gallong",
           titlefont=list(family="Modern Computer Roman"))
```

## New Plot
```{r, message=FALSE, echo=TRUE}
plot_ly(mtcars, x=wt, y=mpg, type="scatter", mode="markers", col=disp) %>%
  layout(xaxis=xax, yaxis=yax)
```

## color, cylinder
```{r, message=FALSE, echo=FALSE}
data(mtcars)
attach(mtcars)
plot_ly(mtcars, x=wt, y=mpg, mode="markers",
        color=as.factor(cyl), size=hp,
        width=600, height=350) %>%
  layout(xaxis=xax, yaxis=yax)
```

## Equation
<p style="font-size:20px">
$X \sim N(\mu=30, \sigma^2=5^2); Y \sim N(0,1); Z \sim N(15, 2^2); color = Z$ <br>
$X, Y, Z - independent$ 
</p>

```{r, message=FALSE, echo=TRUE}
set.seed(123)
myX = rnorm(100, mean=30, sd=5)
myY = rnorm(100)
myZ = rnorm(100, mean=15, sd=2)
```

## Scatter Plot {.flexbox .vcenter}
```{r}
plot_ly(x=myX, y=myY, z=myZ, 
        type="scatter3d", mode="markers",
        color=myZ) %>%
  hide_colorbar()
```


## Surface and Plane Plots
<p style="font-size:15px">
$f(x,y)=sin(x+1)+y^2$ and $f(x,y)=0.6x + 8$
</p>
```{r, message=FALSE}
x = seq(-6, 6, 0.5)
y = seq(-2, 2, 0.5)
X = matrix(rep(x, length(y)), nrow=length(y), byrow=T)
Y = matrix(rep(y, length(x)), ncol=length(x), byrow=F)
Z = sin(X+1) + Y^2
Zp = 0.6*X + 0.8
fig = plot_ly(x=X, y=Y, z=Z, showscale=F, width=800, height=550) %>%
  add_surface(contours=list(
    z=list(show=T, 
           usecolormap=T, 
           highlightcolor="#ff0000", 
           project=list(z=T)
           )
    )
    )
```


## Plot
```{r}
fig = fig %>% layout(list(
  scene=list(camera=list(eye=list(x=1.87, y=0.88, z=-0.6)))
)) %>%
  add_surface(z=Zp)
fig
```

## Latex
$\cdot, x^2, x^{22}, x_a^{22}, x_{ab}^{22}$
$\bar{x}, \bar{x}_n, \hat{\Theta}$
$\alpha, \beta, \gamma, \epsilon, \varepsilon, \phi, \varphi, \chi^2$
$\displaystyle MSE = {SSE\over n-2} \sum_{i=1}^n (y_i - \hat{y}_i)^2$
$Hello \hspace{3cm} There$

## DAT 301
### Ioslides
#### Overleaf
- png
- jpg
1. Hello
2. There
<br>
*Are we in trouble now?*
**Why Aye Man**

## Last plot
```{r, fig.align='center', fig.cap="Scatter plot"}
plot(mtcars$mpg, mtcars$cyl)
```