---
title: "Lab 2"
author: "Vincent Latona"
date: "2023-02-24"
output: pdf_document
---

```{r setup, include=F}
#Include libraries
library(ggplot2movies)
library(ggplot2)
library(plotly)
library(dplyr)
library(tidyr)
library(tidyverse)

data(movies) #Load and attach data
attach(movies)
df = movies
```

# 1. Years of Production
```{r years, include=T}
youngest = max(df$year) #Youngest - Used inline
oldest = min(df$year) #Oldest - Used inline
range = youngest-oldest #Range - Used inline
```
The oldest movie was produced in `r oldest` and the youngest movie was produced in `r youngest`. \newline
The range of movies is `r range` years.

# 2. Movie Budgets
```{r budgets, include=T, comment=""}
total = nrow(df) #Get total number of rows
noBudget = sum(is.na(df$budget)) #Count movies without budget
noBudgetProp = round(noBudget/total, 2) #Calculate proportion - Used inline
budgetProp = round(1-noBudgetProp, 2) #Calculate opposite proportion - Used inline
top = df %>% select(title, budget) #Select needed columns
top = top[order(-top$budget),] %>% head(5) #Order by budget and keep top 5
print.data.frame(top) #Format data frame display
```
The 5 most expensive movies in the data frame are listed above. \newline
The proportion of movies that have their budgets included in the data frame is `r budgetProp` and the proportion of movies that do not have their budgets included in the data frame is `r noBudgetProp`. 
\newpage

# 3. Longest Movies
```{r longest, include=T, comment=""}
long = df %>% select(title, length) #Select needed columns
long = long[order(-long$length),] %>% head(5) #Order by length and keep top 5
print.data.frame(long) #Format data frame display
```
The 5 longest movies in the data frame are listed above.\newline

# 4. Short Movies
```{r short, include=T}
shortMovies = df %>% filter(Short == 1) %>%  #Filter by short
  select(title, length) #Project title and length
shortest = shortMovies[order(shortMovies$length),] %>% head(1) #Get shortest - Used inline
longest = shortMovies[order(shortMovies$length),] %>% tail(1) #Get longest - Used inline
```
The shortest short movie is "`r shortest$title[1]`" with a run time of `r shortest$length[1]` minute(s). \newline
The longest short movie is "`r longest$title[1]`" with a run time of `r longest$length[1]` minute(s). 
\newpage

# 5. Movies per Genre
```{r genres, include=T}
genres = data.frame(Count = c(sum(df$Action), sum(df$Animation), #Define count per genre
                              sum(df$Comedy), sum(df$Drama), 
                              sum(df$Documentary), sum(df$Romance), 
                              sum(df$Short)),
                    Genre = c("Action", "Animation", #Define genres
                              "Comedy", "Drama", "Documentary",
                              "Romance", "Short"))
ggplot(genres, aes(x=Genre, y=Count, fill=Genre)) + #Generate ggplot
  labs(title="Movies per Genre", x="Genre", y="Number of Movies") + #Add labels
  geom_bar(stat="identity") + #Use bar plot
  geom_text(aes(label=Count), vjust=-0.5) #Add values
```
\newpage

# 6. Average Rating per Genre
```{r ratings, include=T}
genreRatings = list(df$rating[df$Action == 1], df$rating[df$Animation == 1], #Calculate ratings
              df$rating[df$Comedy == 1], df$rating[df$Drama == 1],           #per genre
              df$rating[df$Documentary == 1], df$rating[df$Romance == 1],
              df$rating[df$Short == 1])
meanRatings = round(sapply(genreRatings, mean), 4) #Calculate avg ratings
ratingsByGenre = data.frame(Average = meanRatings, #Create data frame
                            Genre = c("Action", "Animation", #Define genres
                              "Comedy", "Drama", "Documentary",
                              "Romance", "Short"))
ggplot(ratingsByGenre, aes(x=Genre, y=Average, fill=Genre)) + #Generate ggplot
  labs(title="Average Rating per Genre", #Add labels
       x="Genre", y="Average Rating of Movies") + 
  geom_bar(stat="identity") + #Use bar plot
  geom_text(aes(label=Average), vjust=-0.5) #Add values
```
\newpage

# 7. Average Rating of Movies from 2000-2005
```{r select rating, include=T}
range = df %>% filter(year %in% 2000:2005) #Select movies from 2000-2005
genreRatings = list(range$rating[range$Action == 1], #Calculate ratings
                    range$rating[range$Animation == 1], #per genre
                    range$rating[range$Comedy == 1], range$rating[range$Drama == 1],
                    range$rating[range$Documentary == 1], range$rating[range$Romance == 1],
                    range$rating[range$Short == 1])
meanRatings = round(sapply(genreRatings, mean), 4) #Calculate avg ratings
ratingsByGenre = data.frame(Average = meanRatings, #Create data frame
                            Genre = c("Action", "Animation", #Define genres
                              "Comedy", "Drama", "Documentary",
                              "Romance", "Short"))
ggplot(ratingsByGenre, aes(x=Genre, y=Average, fill=Genre)) + #Generate ggplot
  labs(title="Average Rating per Genre from 2000-2005", #Add labels
       x="Genre", y="Average Rating of Movies") + 
  geom_bar(stat="identity") + #Use bar plot
  geom_text(aes(label=Average), vjust=-0.5) #Add values
```
\newpage

# 8. Movies by Genre from 1990-2005
```{r name, include=T}
range = 1990:max(df$year) #Specify range of years to select
movieRange = df %>% filter(year %in% range) #Select movies in the range
action = as.vector(tapply(movieRange$Action, movieRange$year, sum)) #Calculate count per
animation = as.vector(tapply(movieRange$Animation, movieRange$year, sum)) #year by genre
comedy = as.vector(tapply(movieRange$Comedy, movieRange$year, sum))
drama = as.vector(tapply(movieRange$Drama, movieRange$year, sum))
documentary = as.vector(tapply(movieRange$Documentary, movieRange$year, sum))
romance = as.vector(tapply(movieRange$Romance, movieRange$year, sum))
genreByYear = data.frame(Year=range, #Generate data frame
                         Action=action, Animation=animation,
                         Comedy=comedy, Drama=drama, 
                         Documentary=documentary, 
                         Romance=romance)
g = gather(genreByYear, key="Genre", value="Count", 2:7) #Gather data by genre
ggplot(g, aes(x=Year, y=Count, color=Genre)) + #Generate ggplot
  geom_line() + #Add lines per genre
  labs(title="Movies by Genre from 1990-2005", #Add labels
       y="Movie Count")
```
\newpage

# 9. Questions relating to Comedy
```{r old men, include=T}
gom = df$rating[df$title == "Grumpy Old Men"] #Used inline
gom2 = df$rating[df$title == "Grumpier Old Men"] #Used inline
best = df %>% 
  filter(Comedy == 1) %>%
  select(rating, title)
best = best[order(-best$rating),]
comedies = df %>% #Filter by comedies in 1993
  filter(Comedy == 1 & year == 1993)
com = c(round(mean(comedies$rating),1), #Define ratings comparison
        comedies$rating[comedies$title == "Grumpy Old Men"])
compare = data.frame(Ratings=com, #Put ratings in data frame
                     Movies=c("Other Comedies",
                              "Grumpy Old Men"))
ggplot(compare, aes(x=Movies, y=Ratings, fill=Movies)) + #Generate ggplot
  geom_bar(stat="identity") + #Use bar plot
  labs(title="Grumpy Old Men vs Average Comedy in 1993") + #Add title
  geom_text(aes(label=Ratings), vjust=-0.5) #Add values
```
1. How different is the user rating for "Grumpy Old Men" compared against the average of user ratings for all comedies in 1993? (answer above) \newline
2. Which film was rated more favorably, "Grumpy Old Men" or "Grumpier Old Men"? \newline
(Grumpy Old Men: `r gom` > Grumpier Old Men: `r gom2`) \newline
3. What movie is the best rated comedy in the database? \newline
(The best rated comedy in the database is "`r best$title[1]`")