---
title: "Lab 1"
author: "Vincent Latona"
date: "2023-02-09"
output: pdf_document
---

```{r, setup, include=FALSE}
getwd()
df = read.csv("Absenteeism_at_work.csv", sep=";", header=TRUE)
library(dplyr)
```

# Height vs Weight

```{r, scatter plot, include=TRUE, echo=TRUE}
x = df$Weight #Assign weight to x-axis
y = df$Height #Assign height to y-axis
plot(x, y, col="navyblue", pch=19 , #Set data to plot, color, and point style
     xlab="Weight of Absentee (kg)", ylab="Height of Absentee (cm)", #Set labels
     main="Scatter Plot of Height vs Weight for Absentees") #Set title
```

-   The information displayed by the graph does not suggest any discernible patterns between height and weight of the data points. \newpage

# Hours of Absences

```{r, hours histogram, include=TRUE, echo=TRUE}
hours = df$Absenteeism.time.in.hours #Store hours
hist(hours, breaks=sqrt(length(hours)), col="red", #Set histogram information
     xlim=c(0,125), ylim=c(0,500), #Set limits
     xlab="Absence Hours", ylab="Number of Absences", #Set labels
     main="Histogram of Absence Hours") #Set title
```

-   The information displayed by the histogram indicates that the majority of the absences are less than 20 hours per absence. Most absences fall between 1 and 10 hours. \newpage

# Age vs Absences

```{r, age histogram, include=TRUE, echo=TRUE}
age = df$Age #Store age
hist(age, breaks=sqrt(length(age)), col="darkorange", #Set histogram information
     xlim=c(min(age)-1, max(age)+1), ylim=c(0,150), #Set limits
     xlab="Age of Absentees", ylab="Number of Absences", #Set labels
     main="Histogram of Age of Absentees") #Set title
```

-   The information displayed by the histogram indicates that there are more absences among younger workers (less than 40 years old). \newpage

# Absence Hours by Month

```{r, monthly hours, include=TRUE, echo=TRUE}
hrspmTibble = group_by(df, Month.of.absence) %>% #Group by month
  summarise(Absenteeism.time.in.hours = sum(Absenteeism.time.in.hours)) #Summarize hours by month
hrs = hrspmTibble$Absenteeism.time.in.hours[-c(1)] #Create vector of hours by month
barplot(hrs, col=topo.colors(length(hrs)), #Set bar plot information
        main="Absence Hours per Month", ylim=c(0, 800), #Set title and limits
        names.arg=c("Jan", "Feb", "Mar", "Apr", "May", "Jun", #Label bars
                   "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"),
        cex.names=0.75, xlab="Months", ylab="Total Hours") #Set labels
```

-   The information provided by the bar plot indicates high volume of absences in July and March. \newline
-   Absences appear to be slightly higher towards the end of the year, possibly because of holidays. \newpage

# Social Smokers

```{r, social smokers, include=TRUE, echo=TRUE}
status = recode_factor(df$Social.smoker, '0'="NSM",'1'="SM") #Encode smoker status
boxplot(df$Absenteeism.time.in.hours ~ status, horizontal=T, #Set box plot information
        ylab="Smoker Status", xlab="Absenteeism in Hours", #Set labels
        main="Absenteeism vs Social Smoker Status", #Set title
        col=c("green", "red")) #Set box colors
legend("topright", legend=c("Non-Social Smoker", "Social Smoker"), #Create legend
       fill=c("green", "red"))
```

-   The information provided by the box plots does not necessarily indicate that being a social smoker influences hours of absenteeism. There are more outliers in the non-social smokers category. \newpage

# Social Drinkers

```{r, social drinkers, include=TRUE, echo=TRUE}
drinkStatus = recode_factor(df$Social.drinker, #Encode drinker status
                            '0'="NSD", '1'="SD")
boxplot(df$Absenteeism.time.in.hours ~ drinkStatus, horizontal=T, #Set bar plot information
        ylab="Drinker Status", xlab="Absenteeism in Hours", #Set labels
        main="Absenteeism vs Social Drinker Status", #Set titles
        col=c("green", "red")) #Set box colors
legend("topright", legend=c("Non-Social Drinker", "Social Drinker"), #Create legend
       fill=c("green", "red"))
```

-   The information provided by the box plots are very similar in distribution and both have a high number of outliers. The information presented by the box plots does not necessarily indicate that social drinker status influences hours of absenteeism.
